%% thesis.tex 2014/04/11
%
% Based on sample files of unknown authorship.
%
% The Current Maintainer of this work is Paul Vojta.

\documentclass{ucbthesis}

\usepackage[american]{babel}
\usepackage{csquotes}
\usepackage[backend=biber, style=apa]{biblatex}
\DeclareLanguageMapping{american}{american-apa}
\bibliography{../../Documents/References/references.bib}

\usepackage{amsmath}


% To compile this file, run "latex thesis", then "biber thesis"
% (or "bibtex thesis", if the output from latex asks for that instead),
% and then "latex thesis" (without the quotes in each case).

% Double spacing, if you want it.  Do not use for the final copy.
% \def\dsp{\def\baselinestretch{2.0}\large\normalsize}
% \dsp

% If the Grad. Division insists that the first paragraph of a section
% be indented (like the others), then include this line:
% \usepackage{indentfirst}

\newtheorem{theorem}{Jibberish}

\hyphenation{mar-gin-al-ia}
\hyphenation{bra-va-do}

\begin{document}

% Declarations for Front Matter

\title{Bayesian and frequentist cross-validation methods for clustered and cross-clustered data}
\author{Daniel C. Furr}
\degreesemester{Spring}
\degreeyear{1995}
\degree{Doctor of Philosophy}
\chair{Professor Sophia Rabe-Hesketh}
\othermembers{Professor Alan Hubbard \\
              Assistant Professor Zacahary Pardos}
\numberofmembers{3}
\field{Education}
\campus{Berkeley}


\maketitle
% Delete (or comment out) the \approvalpage line for the final version.
\approvalpage
\copyrightpage

\include{abstract}

\begin{frontmatter}

\begin{dedication}
\null\vfil
\begin{center}
To Ossie Bernosky\\\vspace{12pt}
And exposition? Of go. No upstairs do fingering. Or obstructive, or purposeful.
In the glitter. For so talented. Which is confines cocoa accomplished.
Masterpiece as devoted. My primal the narcotic. For cine? To by recollection
bleeding. That calf are infant. In clause. Be a popularly. A as midnight
transcript alike. Washable an acre. To canned, silence in foreign.
\end{center}
\vfil\null
\end{dedication}

% You can delete the \clearpage lines if you don't want these to start on
% separate pages.

\tableofcontents
\clearpage
\listoffigures
\clearpage
\listoftables

\begin{acknowledgements}
Bovinely invasive brag; cerulean forebearance.
Washable an acre. To canned, silence in foreign.
Be a popularly. A as midnight transcript alike.
To by recollection bleeding. That calf are infant. In clause.
Buckaroo loquaciousness?  Aristotelian!
Masterpiece as devoted. My primal the narcotic. For cine?
In the glitter. For so talented. Which is confines cocoa accomplished.
Or obstructive, or purposeful.
And exposition? Of go. No upstairs do fingering.

\end{acknowledgements}

\end{frontmatter}

\pagestyle{headings}

% (Optional) \part{First Part}

\newpage
\chapter{Cross-validation for cross-clustered data}
\input{chapter_1}

\newpage
\chapter{Frequentist cross-validation with the doubly explanatory model}
\chaptermark{Frequentist cross-validation}
\input{chapter_2}

\newpage
\chapter{Bayesian cross-validation with the doubly explanatory model}
\chaptermark{Bayesian cross-validation}
\input{chapter_3}


% \appendix
% \chapter{More Monticello Candidates}

\printbibliography

\end{document}
